---
title: "Movie ratings project report"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# By Ilya Letnik

### Introduction:

The goal of the project is to create a movie recommendation system that uses machine learning algorithm. 
The training data used for the analysis is a subset of a larger movie ratings dataset.
The key steps to achieve the goal were: Create the dataset, the test and validation sets, create the basic algorithm, adjust the parameters according to data until RMSE is sufficiently low. 

### Method

1. The subset database was created by taking relevant columns from 2 data files, the validation set will consist of 10% of the data.

2. Account for movie effect (because some movies are rated generally higher) by considering average movie rating in the model

3. Account for user effect (i.e. some users rate all movies high) by considering average user rating in the model



```{r,echo= FALSE, include=FALSE}

#############################################################
# Create edx set, validation set, and submission file
#############################################################

# Note: this process could take a couple of minutes

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- read.table(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                      col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
                                           title = as.character(title),
                                           genres = as.character(genres))

movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data

set.seed(1)
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set

validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set

removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)


#### My Code ###
#account for Movie effect
library(caret)
set.seed(1)

mu <- mean(edx$rating) 
movie_avgs <- edx %>% 
  group_by(movieId) %>% 
  summarize(b_i = mean(rating - mu))

predicted_ratings <- mu + validation %>% 
  left_join(movie_avgs, by='movieId') %>%
  pull(b_i)

model_1_rmse <- RMSE(predicted_ratings, validation$rating)

#account for user effect
user_avgs <- edx %>% 
  left_join(movie_avgs, by='movieId') %>%
  group_by(userId) %>%
  summarize(b_u = mean(rating - mu - b_i))

predicted_ratings <- validation %>% 
  left_join(movie_avgs, by='movieId') %>%
  left_join(user_avgs, by='userId') %>%
  mutate(pred = mu + b_i + b_u) %>%
  pull(pred)
```

### Results 
It was found that using this method design the RMSE could be lowered to the desired level of RMSE <= 0.87750

```{r, echo=FALSE}
model_2_rmse <- RMSE(predicted_ratings, validation$rating)
print("The RMSE for the build model is:")
model_2_rmse

```


### Conclusion
It is possible to design a good movie rating prediction system using a relatively simple machine learning algorithm as shown here.

